### 2018-06-11 express2018/part2.tex
Stage work on part2

\section{Formalization}

In the second half of this paper, we present a formalization of CCS
including the new concepts and theorems proposed in the first half of
this paper. The main purpose is to convince the readers that, there's no flaw in the informal proofs. The
formalization work is based on a formalization of CCS in HOL theorem
prover (HOL4), described in second author's MSc thesis
\cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
The work is essentially a porting and extension of the early CCS
formalization by Monica Nesi \cite{Nesi} in HOL88.

We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of the hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

\subsection{Simplifications and Limitations}

We have only formally verified simplified forms of the unique
solution theorems introduced in the first part. This is to not only
save the formalization efforts, but also to make us focusing on the most
important part of the proof. The most important simplification is
that, we have only proved all unique solution theorems with only single-variable equations
(contractions).

But such a simplification doesn't hurt the rigorousness of our entire
work, as once the single-equation version of the theorem is
proven, the general case is just a routine
adaptation for paper proofs. \footnote{However, it's impossible to directly derive the
  multi-variable versions of unique solution theorems from their
  single-variable versions.} \footnote{Also, I don't know if there
  exists a higher-level general
  duality theorem such that some kind of results on single-variable CCS
  equations must also hold for multi-variable equation groups. --Chun.}
It's not easy to formalize multi-variable CCS equations with
introducing a large number of definitions and intermediate
results. It's easy, however, to directly use the semantics context to
represent single-variable equations without the need of formalizing
the concept of ``equation'' at all.

There're two major limitations in this work. One is the class of CCS
processes, i.e. we're limited with Finitary CCS which has no infinite
sums or parallel compositions. The other limitation is, even for single-variable equation, the
variable doesn't appear inside any constant.

@@

In our formalization of CCS, the set of transition actions are not
defined by the union of all visiable actions (names and conames) and the invisible
action $\tau$. Instead, we define only visible names as a simple
algebraic datatype ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}`` in which the type parameter
$\beta$ can be any other type (thus we don't know anything about the
cardinality of the set of all visiable actions), and then use HOL's
\texttt{optionTheory} to further extend the type by one more
element, $\tau$. Having a dedicated type for all visiable actions is quite
convenient and make some statements shorter. (e.g. the restriction
operator can only take a set of visible actions)

The type of CCS process is defined by an algebraic datatype in HOL,
i.e. an inductive defined finite structure constructed by unary and
binary operators.\footnote{It seems possible to switch to co-algebraic
datatypes for the support of potential infinite sums. Indeed, we never
need to induct on CCS processes in any proof so far.} 
It's not easy to directly extend this datatype with new operators,
without hurting the existing proved results.\footnote{This is usually
  called \emph{deep embedding} when implementing a new logic in theorem
  provers. In case of CCS, \emph{shallow embedding} seems impossible.}
Relabeling operator is not necessary from the theoretical view, as
it's possible to use \emph{syntactic
  substitution} instead\footnote{c.f. Section 4.1.2. (page 168) of Robert Gorrieri's CCS book.}
However, it's quite hard to formalize syntactic substitution for CCS,
as it requires to prove many results on the free and bound names
first.

Without constants (or recursion operator), a CCS process cannot
represent infinite-state LTS and many deep results in CCS cease to
exist then. Therefore any serious CCS formalization
must contain the support of constants (or recursion operator. Nowadays
most CCS textbooks and papers use constants, but from the formal view
it's better (or only possible) to use the original form: the recursion
operator, because any CCS process must be presented as single, finite term in
theorem prover. In Milner's book, this CCS operator is called \texttt{fix},
as it essentially returns the fixpoint of a CCS equation. We use the
symbol \texttt{rec} instead and limit ourself to single-variable case,
as we want to extend to multi-variable recursion operator in the
future with the original name. So instead of saying a CCS process $A$
with constant definitions
$A \overset{\mathrm{def}}{=} \tau.A + b.B$ and $B
\overset{\mathrm{def}}{=} b.B$, we use $\mathbf{rec}\;A.\, (\tau.A +
b.(\mathbf{rec}\;B.\,(b.B)))$ instead.

The extra beauty of above recursion opeator is, if we imagine the
\texttt{rec} symbol as $\lambda$ in $\lambda$-expressions, i.e. $\lambda A.\, (\tau.A +
b.(\lambda B.\,(b.B)))$, then we can see both $A$ and $B$ are bound
variables in this expression. On the other side, it's easy to write
an extra constant $C$ inside previous process without a surrounding $\lambda C$, and it must
be a \emph{free variable} of that $\lambda$-expressions. If we go back to the constant-based CCS
syntax, this would mean a malformed CCS process because not all the
constants used in it are defined. Our point here is:
\begin{enumerate}
\item Under the recursion-based CCS syntax, all possible CCS terms
  constructed from the algebraic datatype are well-formed, and those
  free variables can be seen and used\footnote{However, in our
    formalization of unique solution of equations theorems, we focus
    on single-variable case and directly used $\lambda$-expression
    (typed: $CCS\rightarrow CCS$) as equation. The current discussion
    is mainly for the support of multi-variable equations in the future.} as equation variables.
\item All proved theorems about CCS naturally take CCS process with
  extra free variables, as they by themselves have no transition, just
  like $\mathrm{0}$. (this is a natural consequence derived from SOS rules)
\end{enumerate}

The transition of CCS processes is defined by an inductive
relation, literally the same as SOS rules appearing in CCS
textbooks.\footnote{There's an interesting finding here: we never need to use the induction
theorem for transition relation at all, after having proved almost all
notable results for CCS. In another word, the whole CCS formal theory
builds normally, even if the transition relation
is defined co-inductively (by the same set of SOS rules). A potential
need is the prove the set of ending process of transitions from any
process doesn't use any new labels, but so far this result is not
formalized yet. The question
is: do the largest fixpoint and smallest fixpoint \emph{coincide} for
a relation defined by SOS rules? (if not, what else do we have in the
co-inductive relation?)} The only notable SOS rule which is different
with their usual version in textbooks, is the recursion rule:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
\end{alltt}
which says, if we substitute all appearences of constant $X$ in $E$ to
$(\mathbf{rec}\; X E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; X E)$ has the same
transition.

@@

The major difficultis in formalizing Sangiorgi's unique solution of
contractions theorem is to prove its main lemma (c.f. Lemma 3.9 of
main paper), in which the length of weak transitions are used for
induction. Introducing a special vesion of weak transition relation
with length would be an easier choice, but we have chosen to use trace
instead, so that in the future it's possible to extend the work with
trace equivalence included.

>>>>

@@

In the formalization of CCS, it's better to have two distinct
types for actions: the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Label}} ($\beta$ is a type variable.) for visible actions, divided by input
and output actions:
\begin{lstlisting}
val _ = Datatype `Label = name 'b | coname 'b`;
\end{lstlisting}
and the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Action}} as the
union of all visible and invisible actions. The cardinality of all
available types depends on the choice of type variable $\beta$.

The core datatype ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS}}'' is then defined as an inductive
datatype in HOL, based on its Datatype package:
\begin{lstlisting}
val _ = Datatype `CCS = nil
		      | var 'a
		      | prefix ('b Action) CCS
		      | sum CCS CCS
		      | par CCS CCS
		      | restr (('b Label) set) CCS
		      | relab CCS ('b Relabeling)
		      | rec 'a CCS`;
\end{lstlisting}

We have added some minimal grammar support using HOL's powerful pretty printer, to represent CCS
processes in more readable forms. Table \ref{tab:ccsoperator} has listed the notation of typical CCS processes and
major operators supported by above definition. (Notice the use of
recursition operator for representing process constants)
\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operator} & \textbf{Notation} & \textbf{HOL} & \textbf{HOL
  (alternative)}\\
\hline
nil & $\textbf{0}$ &  \HOLinline{\HOLConst{nil}} & \HOLinline{\HOLConst{nil}} \\
Rrefix & $a.b.0$ & \texttt{prefix a (prefix b nil)} & \HOLinline{\HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{b}\HOLSymConst{..}\HOLConst{nil}} \\
Sum & $p + q$ & \texttt{sum p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{+} \HOLFreeVar{q}} \\
Parallel & $p \,\mid\, q$ & \texttt{par p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{q}} \\
Restriction & $(\nu\;L)\;p$ & \texttt{restr L p} & \HOLinline{\HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{p}} \\
Constant & $A=a.A$ & \texttt{rec A (prefix a (var A))} & \HOLinline{\HOLConst{rec} \HOLFreeVar{A} (\HOLFreeVar{a}\HOLSymConst{..}\HOLConst{var} \HOLFreeVar{A})} \\
\hline
\end{tabular}
\end{center}
   \caption{Syntax of CCS operators}
   \label{tab:ccsoperator}
\end{table}

The transition semantics of CCS processes were defined by the following
Structural Operational Semantics (SOS for short) rules in
Fig.\;\ref{f:LTSCCS}, except that the transition rule for constants
has been changed to recursion operator:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
\end{alltt}
which says, if we substitute all appearences of constant $X$ in $E$ to
$(\mathbf{rec}\; X\; E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; X\; E)$ has the same
transition.

@@

The major difficulties in formalizing Sangiorgi's unique solution of
contractions theorem is to prove its main lemma (c.f. Lemma 3.9 of
main paper), in which the length of weak transitions are used for
induction. Introducing a special vesion of weak transition relation
with length would be an easier choice, but we have chosen to use trace
instead, so that in the future it's possible to extend the work with
trace equivalence included.
